% !TEX TS-program = lualatexmk
% !TEX parameter =  --shell-escape

\documentclass[11pt]{book}

\input{preamble_l_e_book}

\begin{document}

\gscore[]{}{arc_no_text}

\newpage

\smallscore{arc_solfege}

\newpage

\emph{The book gives these chants as examples to solfège. It's a good exercise, so why not learn the chants in full? They can be sung at communion, the first two for exposing the Blessed Sacrament (instead of \emph{O Salutaris}) and the latter for the Marian chant: use the same versicle, response, and collect as is done in the preconciliar office of Compline, as we also do after the high Mass at Assumption.}

\gscore[Ant.]{6.}{an_ave_regina_caelorum_simple_tone_solesmes}

Hail, O Queen of Heaven.
Hail, O Lady of Angels
Hail! thou root, hail! thou gate
From whom unto the world a light has arisen:
Rejoice, O glorious Virgin,
Lovely beyond all others,
Farewell, o exceedingly beautiful maiden,
And pray for us to Christ.

\gscore[]{6}{va_ave_verum_solesmes}

Hail, true Body, born
of the Virgin Mary,
having truly suffered, sacrificed
on the cross for mankind,
from whose pierced side
water and blood flowed:
Be for us a foretaste [of the Heavenly banquet]
in the trial of death! O sweet Jesus, O holy Jesus,
O Jesus, son of Mary.

\gscore[]{5.}{hy_adoro_te_devote_solesmes}

I devoutly adore thee, hidden deity,
Who are truly hidden beneath these appearances.
My whole heart submits to thee,
because in contemplating thee, it is fully deficient.

Sight, touch, taste all fail in their judgment of thee,
But hearing suffices firmly to believe.
I believe all that the Son of God has spoken;
There is nothing truer than this word of truth.

On the cross only the divinity was hidden,
But here the humanity is also hidden.
Yet believing and confessing both,
I ask for what the penitent thief asked.

I do not see wounds as Thomas did,
But I confess that thou art my God.
Make me believe much more in thee,
Hope in thee, and love thee.

O memorial of our Lord's death,
Living Bread that gives life to man,
Grant my soul to live on thee,
And always to savor thy sweetness.

Lord Jesus, Good Pelican,
clean me, the unclean, with thy Blood,
One drop of which can heal
the entire world of all its sins.

Jesus, whom now I see hidden,
I ask thee to fulfill what I so desire:
That the sight of thy Face being unveiled
I may have the happiness of seeing thy glory. Amen.

\gscore[]{6}{va_inviolata_solesmes}

Inviolate, whole and chaste art thou, Mary:
Thou art the shining gate of heaven.
O kind mother, dearest to Christ,
receive our pious proclamations of praise.
To thee our devout hearts and mouths cry out:
may our souls and bodies be pure.
Through thy sweet sounding prayers
may thou grant us forgiveness forever.
O kindly one! O Queen! O Mary!
Thou alone didst remain inviolate.

\enddocument